Memories of migrations past: Sociality and cognition in dynamic, seasonal environments

### 2021-07-16 text/SocialMemoryMigration_FEE.tex
...

Proximate causes, drivers and mechanism can vary considerably across and even within species \citep{Berthold1999, Shaw2016}. Some migrants follow a ``green wave'' of spring vegetation as it flowers across altitudinal or latitudinal gradients \citep{Bischof2012, Kolzsch2015, Merkle2016}. These migrations can be considered ``tactical'', as they can occur - as an extreme simplification - purely as response to local conditions. Other migrants perform long-distance migrations in anticipation that critical resources will be available at the time of arrival at the end point of migration \citep{Abrahms2019}. This second behavior involves the greatest trade-off between the costs of migration against the benefits of accessing resources, whether those are food, suitable habitat for breeding, or refuge from predators, that are highly seasonal and localized. This approach can be considered ``strategic'' in the sense that it is driven not by immediate cues but by an anticipation based on prior experience \citep{Bracis2017}.

Whether long distance migration is a Migratory species are often considered to be more vulnerable to environmental change as disruption in either of the seasonal ranges or along a migratory corridor can have significant negative impacts \citep{Wilcove2008, Kauffman2021}. On the other hand, migratory species might be more resilient to large-scale disruptions due to their wide-ranging mobility \citep{Robinson2009}. Clearly, the ability to perform a migration without local cues is only possible if the behavior is hard-programmed or remembered. On the other hand, a strictly programmed behavior can be maladaptive if conditions change. Because the scenarios underlying migration are manifold and complex, mathematical modeling may provide some insights and help clarify where, when, and under what conditions we might migration behavior to emerge, to be adaptive, to be maladaptive, or to collapse.

Here, we develop a social diffusion-advection model with memory to explore the resilience of a migratory population under various dynamic, seasonal resource distributions. In formulating the model, our goal was to identify the minimum set of movement and memory parameters required to generate an adaptive, migratory behavior. This includes the ability to learn to migrate from non-migratory initial conditions, simulating the release of naive animals in a seasonal environment \citep{Jesmer2018}, to lose the propensity to migrate if the resource distribution does not require it, also a commonly observed phenomenon \citep{Wilcove2008}, and to assess the resilience or fragility of a migratory population against changing resource distribution dynamics, including both stochasticity and trends in spatial and temporal distributions, mirroring the effects of climate change \citep{Park2020}.

@@

Animals navigate complex, dynamic, patchy environments. When there is a strongly localized and seasonal component to the resource dynamics, movement strategies limited to straightforward resource-following taxis necessarily fail to maximally exploit the dynamics. It is in these cases - quite common in the natural world -- that seasonal migration becomes a viable, even necessary, strategy. However, when resources start shifting in space and time -- as is occurring at an accelerated pace with recent global climate change -- the migration phenology itself must exhibit some plasticity. It is our conjecture that this plasticity is greatly facilitated by a memory-driven process, in which recent experiences inform strategic behaviors in subsequent years. Here, we proposed a relatively simple, memory-driven mechanism by which a population can respond to long-term changes in resource dynamics, and a framework within which interacting memory, movement, social and resource dynamics can be explored.

We found that the straightforward inclusion of a migration-specific memory in a social diffusion-advection framework was sufficient to capture many fundamental phenomena related to migration. In essence, our model allowed a population with a hybrid migratory and resource-following behavior to adjust the migratory behavior based on recent experiences with the resource location. The model, while simple, was able to emulate the successful navigation of an environment with temporally and spatially isolated seasonal resource patches, the spontaneous emergence of a migratory behavior, and intrinsic robustness to changes in those environmental resources, whether those changes were steadily shifting trends or inter-annual stochasticity. We postulated a process defined by relatively few parameters, and explored the ability of the process to navigate a fixed set of environments. Our goal was to understand the combinations of factors that lead to a resilient migration behavior. The foraging efficiency was a convenient metric of the utility of the migratory behavior, but was not a measure explicitly maximized by the model. Other metrics, such as foraging efficiency in a given season, may respond differently across model parameters and could be useful in understanding the relative success of alternative migratory strategies in different contexts (e.g., scenarios where there are differing constraints between 'low' and 'high' seasons or where good resource matching is particularly vital at certain stages in an animal's life history like gestation).  However, we opted for an overall annually averaged metric in these initial explorations because that provides the broadest linkage between resource dynamics and the animals' locations and was consistent with the smooth and continuous nature of the resource distribution.

Our analyses afford several rather general conclusions. One is that migration in general can be acquired, as has been demonstrated for translocated ungulates \citep{Jesmer2018}, even in environments where resources are difficult to ``surf'' (figure \ref{fig_learningtomigrate}). However, this requires a very strong attraction to resources, somewhat higher levels of exploratory behavior, and many years, findings that echo empirical observations \citep{Jesmer2018}. On the other hand, migrations can be considered somewhat fragile, in the sense that under certain conditions, e.g.~increasing stochasticity, rapidly shifting resources, or a shift in some of the system parameters, migration can collapse down into a localized non-migratory, residential behavior, at times with only a marginal cost to the foraging efficiency index. This sensitivity may explain why partially migratory populations composed of a mixture of resident and migratory individuals are so common and, apparently, evolutionary stable \citep{Berthold1999, Chapman2011}.

We also found that the ability to maintain, learn and adapt a migration depends strongly on properties of the resource dynamics. In particular, the reinforcement of memory and foraging is strongest when patches are concentrated in time, but relatively large in space. Interestingly, in most stable patterns, the eventual targeted migration arrival time coincided with the \emph{peak}, rather than the beginning, of the resource dynamic. This indicates that the long-distance social migration behavior may be particularly reinforced when the targeted resource is very sudden. This is the case for the rapid green-up that occurs in high latitudes as snow recedes in tandem with extended day lengths leading to an intense green-up period \citep{Park2020} or, for example, when resources are linked to the short-duration early blooming phenology of very particular plants \citep{Post2007}.

The sociality parameters -- in particular, the spatial scale $\lambda$ -- were, somewhat unexpectedly, the most important for determining the resiliency of the process for stabilizing a migratory behavior. In particular, a population very small spatial scale tended to have a much more difficult time locking in to an adaptive migratory pattern, and only in conditions where the social attraction was relatively weak. In order to adapt to any shift in resources, \emph{some} individuals must experience that resource, and the two sociality-related mechanisms to have that scope are either a relatively large spatial scale or a relatively weak social cohesion. On the other hand, overly large spatial scales compromised the ability of the process to track climate change, due to a dilution of the population's ability to concentrate over available resource patches. \emph{Multiple avenues by which 'sociality' (defined as swarm x attraction interplay) can structure migration depending on the relative balance of diffusion and resource following (figure \ref{fig_}).   For example this could tie into the relative balance between individuals exploring the landscape to acquire new information about resources and their opportunities to interact with other individuals and communicate (or attract).}

>>>>

	Proximate causes, drivers and mechanism can vary considerably across and even within species \citep{Berthold1999, Shaw2016}. Some migrants follow a ``green wave'' of spring vegetation as it flowers across altitudinal or latitudinal gradients \citep{Bischof2012, Kolzsch2015, Merkle2016}. These migrations can be considered ``tactical'', as they can occur - as an extreme simplification - purely as response to local conditions. Other migrants perform long-distance migrations in anticipation that critical resources will be available at the time of arrival at the end point of migration \citep{Abrahms2019}. This second behavior involves the greatest trade-off between the costs of migration against the benefits of accessing resources, whether those are food, suitable habitat for breeding, or refuge from predators, that are highly seasonal and localized. This approach can be considered ``strategic'' in the sense that it is driven not by immediate cues but by an anticipation based on prior experience \citep{Bracis2017, Merkle2019, Bauer2020}.

	Migratory species may be more vulnerable to environmental change as disruption in either of the seasonal ranges or along a migratory corridor can have significant negative impacts \citep{Wilcove2008, Kauffman2021}. On the other hand, migratory species might be more resilient to large-scale disruptions due to their wide-ranging mobility \citep{Robinson2009}. A long-distance migration can not rely entirely on local cues, and is therefore only possible if the behavior is hard-programmed or remembered. On the other hand, a strictly programmed behavior can be maladaptive if conditions change.While migration can clearly be a highly successful strategy, the resilience of migratory populations and migratory behavior is an open question and depends on the plasticity and adaptability of a population \citep{Xu2021}.  That plasticity can take multiple forms, reflecting variation in \textit{where}, \textit{when} and \textit{whether} the migration occurs \citet{Gurarie2017,Xu2021}.

	Cognitive processes, in particular spatial memory, have been shown and argued to be important mechanisms for the reinforcement and maintenance of migration \citep{Merkle2019, Bauer2020}. Similarly, sociality and social learning are likely essential to maintaining migration \citep{Guttal2020,Fagan2011,Jesmer2018,Berdahl2018}.  However, the interacting role of sociality and spatial memory for the plasticity of migration and the resilience of the behavior when faced with a changing environment is largely unstudied.  Because the scenarios underlying migration are manifold and complex, mathematical modeling may provide some insights and help clarify where, when, and under what conditions we might expect migration behavior to emerge, to be adaptive, to be maladaptive, or to collapse.

	Here, we develop a diffusion-advection model with sociality and memory to explore the resilience of a migratory population under various dynamic, seasonal resource distributions. In formulating the model, our goal was to identify the minimum set of movement and memory parameters required to generate an adaptive, migratory behavior. This includes the ability to learn to migrate from non-migratory initial conditions, simulating the release of naive animals in a seasonal environment \citep{Jesmer2018}, to lose the propensity to migrate if the resource distribution does not require it, also a commonly observed phenomenon \citep{Wilcove2008}, and to assess the resilience or fragility of a migratory population against changing resource distribution dynamics, including both stochasticity and trends in spatial and temporal distributions, mirroring the effects of climate change \citep{Park2020}.

@@

	Animals navigate complex, dynamic and patchy environments. When there is a strongly localized and seasonal component to the resource dynamics, movement strategies limited to straightforward resource-following taxis necessarily fail to maximally exploit the dynamics. It is in these cases, quite common in the natural world, that seasonal migration becomes a viable, even necessary, strategy.  However, when resources start shifting in space and time -- as is occurring at an accelerated pace with recent global climate change -- the migration phenology itself must exhibit some plasticity. It is our conjecture that this plasticity is facilitated by a memory-driven process, in which recent experiences inform strategic behaviors in subsequent years, combined with a social dynamic. The relatively simple, social and memory-driven mechanism we  propose here was, in fact, able to adapt to long-term changes in resource dynamics, even with inter-annual stochasticity. The model thereby provides a framework with which the interaction  of memory, movement, social and resource dynamics can be explored.

	We found that the straightforward inclusion of a migration-specific memory in a social diffusion-advection framework was sufficient to capture many fundamental phenomena related to migration. In essence, our model allowed a population with a hybrid migratory and resource-following behavior to adjust the migratory behavior based on recent experiences with the resource location. The model was able to emulate the successful navigation of an environment with temporally and spatially isolated seasonal resource patches, the spontaneous emergence of a migratory behavior, and intrinsic robustness to changes in those environmental resources, whether those changes were steadily shifting trends or inter-annual stochasticity.  In a generalized way, it captured all three fundamental types of migration plasticity: the \textit{where}, the \textit{when} and the \textit{whether} \citep{Xu2021}.

	Our goal was to understand the combinations of factors that lead to a resilient migration behavior. We used foraging efficiency as a convenient metric of the utility of migration, but this was not a measure explicitly maximized by the model. Other metrics, such as foraging efficiency in a given season, or probability of survival or reproduction relative to resource availability \citep{Bauer2020} may respond differently across model parameters and could be useful in understanding the relative success of alternative migratory strategies in different contexts. However, the overall annually averaged metric provided the broadest linkage between resource dynamics and animals' locations and was consistent with the minimal biological assumptions and generality of our framework.

	Among the general conclusions from our model, one is that migration can be acquired in a seasonal environment, even with no strong intrinsic propopensity to migrate and a weak phenological resource pulse to follow. This has been demonstrated for translocated ungulates \citep{Jesmer2018}, even in environments where resources are difficult to ``surf'' (figure \ref{fig_learningtomigrate}). Learning migration, howewer, requires a very strong resource attraction, higher levels of exploratory behavior (e.g.~diffusion, and larger spatial scale of sociliaty), and -- often --- many more years. These findings echo empirical observations \citep{Jesmer2018}.  On the other hand, migrations can be considered somewhat fragile, in the sense that under certain conditions, e.g.~increasing stochasticity, rapidly shifting resources, a shift in some of the system parameters, or even a shift in the spatial and temporal extent of resources, migration can collapse and turn into a non-migratory, residential behavior (figure \ref{fig_phenologymatching}). This sensitivity may explain why partially migratory populations composed of a mixture of resident and migratory individuals are so common and, apparently, evolutionary stable \citep{Berthold1999, Chapman2011}, as well as the wide range of migration plasticity shown in wild populations, even within a species \citep{Xu2021}.

	The ability to maintain, learn and adapt a migration also depends strongly on properties of the resource dynamics. In particular, the reinforcement of memory and foraging is strongest when patches are concentrated in time, but relatively large in space. Interestingly, in most stable patterns, the eventual targeted migration arrival time coincided with the \emph{peak}, rather than the beginning, of the resource dynamic. This indicates that the long-distance social migration behavior may be particularly reinforced when the targeted resource is very sudden. This is the case for the rapid green-up that occurs in high latitudes as snow recedes in tandem with extended day lengths leading to an intense green-up period \citep{Park2020} or, for example, when resources are linked to the short-duration early blooming phenology of very particular plants \citep{Post2007, Renner2018}.

	\textit{The sociality parameters -- in particular, the spatial scale $\lambda$ -- were, unexpectedly, the most important parameters for determining the resiliency of the process. In particular, a population with a small spatial scale tended to have a much more difficult time locking in to an adaptive migratory pattern, and only in conditions where the social attraction was relatively weak. In order to adapt to any shift in resources, \emph{some} individuals must experience that resource, and the two sociality-related mechanisms to have that scope are either a relatively large spatial scale or a relatively weak social cohesion. On the other hand, overly large spatial scales compromised the ability of the process to track climate change, due to a dilution of the population's ability to concentrate over available resource patches. Thus, there appear to be multiple avenues by which 'sociality' can structure migration.   For example this could tie into the relative balance between individuals exploring the landscape to acquire new information about resources and their opportunities to interact with other individuals and communicate (or attract).
	}
